Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-24 analysis/paper/article.qmd
ready to resubmit after peer review

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of undertaking an academic career. Job listing websites often serve as the first port-of-call for students  seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how  archaeological topics, methods, and geographic regions specified in archaeological job ads shifted over time. We also investigate whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students, graduates, and advisors to better understand the dynamics of the academic job market and the requirements of employers.
keywords: |
  archaeology; hiring; academia; job ads 
---

@@

@@

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. Tenure-track positions are a type of academic job that typically involve a probationary period (e.g. 5-7 years) culminating in an evaluation of research, teaching, and service, that if successful, results in permanent employment, job security and academic freedom. This obsessive focus on career trajectories is warranted. As of 2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions [@the20222022]; today, 71% of faculty in the US are non-tenure-track [@culver]. The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors [@beck], including decreasing federal support for higher education, concomitant increases in student debt [@gusterson2017], a pronounced shift in university investment from faculty to administration [@graeber2018, pp. 162–163], and the growing privatization and market orientation of scientific research [@mirowski2011].

@@

Among many scholars, there is also a perception that the last three decades have seen the academic job market become increasingly competitive. Across the social sciences and humanities, there are fewer jobs available relative to the number of PhD graduates, and higher numbers of short-term positions relative to permanent positions[@bessner2021; @bramiPrecariousFutureReflections2023; @kawaSocialNetworkUS2019; @kelsky2015; @mackieMarketShareAccounting2023]. This market saturation has arisen in tandem with increasingly complex application requirements. Many job ads now solicit specific documents in *addition* to the cover letter and CV, such as teaching, research, and diversity statements. These additional documents must be tailored for each application, making the process of applying for jobs a full-time job of its own.

In American anthropology, the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education [@earle2014]. When combined with the institution of tenure, the ADEA exemption allowed faculty to stay in their posts for as long as they liked. The median age for faculty in the U.S. now ranks among the highest for all professions [@kaskie2016]. In tandem with the gradual de-investment in American higher education since 1980s [@mirowski2011] and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, for example, @Passalacqua_2018 found a ratio of 0.81 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, there is one tenure-track position in the US for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. In engineering, Larson et al., have calculated that providing jobs for even 50% of PhD graduates would require the field to expand at an "improbable" rate of 14% per year [-@larson2014, pp.747]. As they emphasize, "... the system in many places is saturated, far beyond capacity to absorb new PhDs in academia at the rates they are being produced" [@larson2014, pp.749].

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. The rising number of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, limited opportunities for professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially due to low compensation combined with expensive and frequent relocation, restrictive socially due to isolation from family and community, and limited professionally due to being ineligible for many decision-making roles at the universities at which they work [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 53 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments and widely cited faculty who hold prestigious awards produce the majority of tenured and tenure-track faculty.

Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of sociology, communication, operations research, and industrial systems engineering show similar dynamics [@barnett2010; @castillo; @feeley2021; @nevin2019]. Broader analyses comparing computer science, business, and history identify comparable patterns among disparate disciplines across the humanities, social sciences, and STEM field—"across disciplines, prestige hierarchies make the most accurate predictions of faculty placement" [@clauset2015, pp. 4]. The most recent comprehensive meta-analysis of faculty placement dynamics in the United States, which examined employment and doctoral education of all tenure-track faculty at PhD-granting universities, underscored the outsize impact of institutional prestige: 80% of faculty trained in the US came from just 20% of institutions [@wapmanQuantifyingHierarchyDynamics2022a]. Within anthropology, such studies draw attention to a central paradox—a discipline purportedly committed to fighting social inequalities continues to reproduce its own systemic inequalities through hiring practices that favor an elite minority of candidates with prestigious affiliations.

@@

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. <!-- **MORE DETAILS ON THIS HERE—JB WILL ADD.** BM: ok, thank you! -->

@@

Collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of editorial activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. The MediaWiki technology used by Academic Jobs Wiki creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics, which allows us to assess the dynamics of editing activity over the ten years of our study period. Every edit includes the author's identity, recorded as either an IP address (a numerical address that includes information about the location of the user's computer) or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the US, indicating the Archaeology Jobs Wiki is primarily edited and visited by users in the U.S. The number of edits and active editors roughly halved from 2013 to 2017, and then continued to gradually decline. Despite this downward trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant across our study period. The majority of editors are active on the site for a single day (n = `r edits_for_each_year_tbl_span_tbl_less_than_one_day`, `r edits_for_each_year_tbl_span_tbl_less_than_one_day_perc`%). There are `r edits_for_each_year_tbl_span_tbl_less_than_one_year` (`r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%) editors who were active for between one day and one year. Only `r edits_for_each_year_tbl_span_tbl_more_than_one_year` (n = `r edits_for_each_year_tbl_span_tbl_more_than_one_year_perc`%) editors were active for more than one year, with `r max(round(edits_for_each_year_tbl_span_tbl$edit_span_years,0))` years as the longest period any single editor was active on the Wiki. For those active for more than one day, the average duration of editorial activity is `r edits_for_each_year_tbl_span_tbl_mean` years. This suggests that most of the editing activity on the Wiki reflects a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads. That said, inferring the number of editors from the number of IP addresses is complicated by the high mobility of job seekers. For example one person could be represented by several IP addresses as they relocate from one city to another while working in a series of short-term appointments such as post-doctoral fellowships and adjunct teaching positions. This might result in an overestimation of the number of editors on the Wiki and underestimation of the duration of their editorial activity.

@@

This trajectory in archaeology—where there appears to be a temporal “sweet spot” post-PhD for obtaining permanent academic positions—dovetails with a larger meta-analysis of the career trajectories of PhDs from the humanities and humanistic social sciences. Main et al. [-@main2019] used longitudinal Graduate Education Survey (GES) data from the Andrew W. Mellon Foundation to analyse the career trajectories of over 5,000 PhDs from disciplines in the humanities and humanistic social sciences at three time points: six months post PhD, three years post PhD, and eight years post-PhD. The proportion of surveyed doctorates in tenure-track or tenured faculty positions increased with time since PhD—41% of the sample held a tenure-track position six months after defending, 64% held a tenure-track position three years after defending, and 68% held a tenure-track or tenured position eight years after defending. In keeping with broader patterns observed across the humanities and social sciences, the variable tenures of editors on the Archaeology Jobs Wiki may thus partially reflect the periodicity of hiring relative to job seekers' time since PhD.

There seems to be a shift over time in the balance of editors who identify as job-seekers versus search committee members, as indicated by the number of self-reporting users over time. Panel G of @fig-panel-wiki-characteristics shows the data found on each page in a section labeled 'current users', where users can volunteer to update a table tracking the status of users. These data indicate that the total number of self-reporting users, from the values recording at 'How many people use the wiki?', declined by `r how_many_people_use_the_wiki_perc_start_end`% during the study period. At the same time, the proportion of users self-identifying as "search committee members" increased, from under `r search_committee_members_perc_start`% in 2014-2018 to over `r search_committee_members_perc_end`% from 2020 onward. This suggests that, over time, fewer job applicants have been contributing job postings and sharing information about their experiences, and search committee members have increasingly been posting job advertisements themselves. Overall, the Archaeology Academic Jobs Wiki is best understood not as a comprehensive archive of all jobs posted during the study period, but a biased sample of the the jobs that early career applicants were most focused on applying for, and sharing information about.

@@

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methodological foci of the position into a Google form. The topical focus is what we understood as the intellectual core of the position—examples of topical foci included environmental archaeology, public archaeology, and North American archaeology. The geographic focus is the region of the world about which the ideal candidate has scholarly expertise, for example, Southwest US, Mediterranean, or Asia and India. The methods focus is the data-generating sub-field of archaeology mentioned in the ad. Examples of methods used in this study include archaeobotany, lithic analysis, and zooarchaeology. We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad.

After completing primary data collection, we studied the topical, geographic, and methods text of each ad. Following the approach of @ryan2003techniques, we collaboratively and manually reduced the variation in the raw data into 10-15 categories appearing in at least 20 (for topics and geography) or 10 (for methods) job ads to simplify analysis and visualization. This means that some topics, such as gender (mentioned in `r n_jobs_mentioning_gender` ads) do not appear in our results because of their rarity in the job ads. Full details of the category reduction, showing the mapping between exact phrases found in the job ads and the categories we used for our analysis, can be found in our Supplementary Materials. Our final topic categories were: American archaeology, Ancient Europe and Mediterranean, Archaeological science, Archaeological theory, Bioarchaeology, Complex societies, Digital archaeology, Environmental archaeology, Evolutionary anthropology, Indigenous and historical archaeology, North Mesoamerican archaeology, Pleistocene archaeology, and Public archaeology. Our geographic categories were: Africa, Americas, Asia & India, Canada & Arctic, Europe, Mediterranean, Meso- & South America, Near East, Oceania, Midwest US, Northeastern US, Southeast US, Southwest US, and Western US. Our methods categories were: Archaeobotany, Archaeometry, Bioarchaeology, Ceramic analysis, Computational and Digital archaeology, Geoarchaeology, Landscape analysis, Lithic analysis, Material culture analysis, and Zooarchaeology.

@@

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx \[ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug\] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities. @fig-show-basic-plots shows the count of ads for each year, where year refers to the year the job ad was posted. @tbl-show-basic-counts shows the breakdown by different job types, ranks, and tenure status. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions. Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

Panel A of @fig-show-map-of-hiring-institution-and-cc shows the frequencies of hiring institutions according to their Carnegie Classification, a framework for classifying US colleges and universities according to the types of degrees awarded, levels of research activity, and topical foci [@shulman2001carnegie]. Doctoral universities with high and very high research activity are by far the most active institutions hiring archaeology faculty. Associate's colleges, also known as community colleges, rarely post job ads for archaeology faculty.

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, Pennsylvania, and Florida. These top five states correspond to the top five most populous US states, suggesting that rates of hiring are approximately proportional to population density. The top five states for job ads are also the five states with highest number of degrees awarded in Anthropology [@IPEDSNCES]. Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.

@@

We recorded all geographic regions mentioned in the text where the successful applicant should have expertise and be research active. Our analysis focuses on those locations mentioned in 20 or more ads. Overall, American locations dominate. Panel A of @fig-show-geo-trends shows that a single region of the US, the Southwest, occurs in more job ads than every other part of the world except for the Mediterranean. The Southwest includes Arizona and New Mexico, with portions of California, Colorado, Nevada, Oklahoma, Texas, and Utah. It is archaeologically significant as the home of the Ancestral Pueblo, Hohokam, and Mogollon peoples, who practiced irrigation agriculture and lived in relatively large settlements compared to other regions of the US. The area was later occupied by the Navajo, Ute, Southern Paiute, Hopi and Zuni, groups who had similarly high population densities [@griffin2000native]. The Mediterranean is prominent because it is the region that is often mentioned in job ads focused on classical archaeology (i.e. archaeology of Bronze Age and Iron Age Italy and Greece).

>>>>

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of undertaking an academic career. Job listing websites often serve as the first port-of-call for students  seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examined changes in the academic job market over time. We investigated the editing dynamics of the Wiki to understand its users and their biases. We then analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how  archaeological topics, methods, and geographic regions specified in archaeological job ads have shifted over time. We also explored whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? Finally, we assessed the influence of socio-political factors on the changing focus of research topics in the field. We conclude with implications for archaeology students, graduates, and advisors seeking to understand the dynamics of the academic job market and the requirements of employers.
keywords: |
  archaeology; hiring; academia; job ads 
---

@@

"The abstract should not exceed 200 words in length."

@@

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. Tenure-track positions are a type of academic job that typically involve a probationary period (e.g. 5–7 years) culminating in an evaluation of research, teaching, and service, that if successful, results in permanent employment, job security and academic freedom. This obsessive focus on career trajectories is warranted. As of 2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions [@the20222022]; today, 71% of faculty in the U.S. are non-tenure-track [@culver]. The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors [@beck], including decreasing federal support for higher education, concomitant increases in student debt [@gusterson2017], a pronounced shift in university investment from faculty to administration [@graeber2018, pp. 162–163], and the growing privatization and market orientation of scientific research [@mirowski2011].

@@

Among many scholars, there is also a perception that the last three decades have seen the academic job market become increasingly competitive. Across the social sciences and humanities, there are fewer jobs available relative to the number of PhD graduates, and higher numbers of short-term positions relative to permanent positions [@bessner2021; @bramiPrecariousFutureReflections2023; @kawaSocialNetworkUS2019; @kelsky2015; @mackieMarketShareAccounting2023]. This market saturation has arisen in tandem with increasingly complex application requirements. Many job ads now solicit specific documents in *addition* to the cover letter and CV, such as teaching, research, and diversity statements. These additional documents must be tailored for each application, making the process of applying for jobs a full-time job of its own.

In American anthropology, the number of doctoral anthropology graduates has increased by about 70% over the past 30 years, but the number of new faculty positions has not increased proportionally [@speakman2018choosing]. New faculty positions have dwindled, in part due to the removal of the of the Age Discrimination in Employment Act (ADEA) exemption in 1994 which prohibited mandatory retirement ages in higher education [@earle2014]. When combined with the institution of tenure, the ADEA exemption allowed faculty to stay in their posts for as long as they liked. The median age for faculty in the U.S. now ranks among the highest for all professions [@kaskie2016]. In tandem with the gradual de-investment in American higher education since 1980s [@mirowski2011] and the aftershocks of the 2008 recession, the ADEA exemption has cultivated an environment where new lines are few and far between. Among biological anthropologists, for example, @Passalacqua_2018 found ratios ranging from 1.21 to 0.82 PhDs to job academic advertisements per year, concluding that academic positions in biological anthropology are barely at sustainable levels. This echoes findings from other fields. In biomedical sciences, there is one tenure-track position in the U.S. for approximately every 6.3 PhD graduates [@Ghaffarzadegan_2015]. In engineering, Larson et al., have calculated that providing jobs for even 50% of PhD graduates would require the field to expand at an "improbable" rate of 14% per year [-@larson2014, pp.747]. As they emphasize, "... the system in many places is saturated, far beyond capacity to absorb new PhDs in academia at the rates they are being produced" [@larson2014, pp.749].

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. The rising number of contingent faculty has long been a concern in U.S. higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in U.S. universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, limited opportunities for professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially due to low compensation combined with expensive and frequent relocation, restrictive socially due to isolation from family and community, and limited professionally due to being ineligible for many decision-making roles at the universities at which they work [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 U.S. graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programs are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the U.S. in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 53 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments and widely cited faculty who hold prestigious awards produce the majority of tenured and tenure-track faculty.

Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of sociology, communication, operations research, and industrial systems engineering show similar dynamics [@barnett2010; @castillo; @feeley2021; @nevin2019]. Broader analyses comparing computer science, business, and history identify comparable patterns among disparate disciplines across the humanities, social sciences, and STEM field—"across disciplines, prestige hierarchies make the most accurate predictions of faculty placement" [@clauset2015, pp. 4]. The most recent comprehensive meta-analysis of faculty placement dynamics in the United States, which examined employment and doctoral education of all tenure-track faculty at PhD-granting universities, underscored the outsize impact of institutional prestige: 80% of faculty trained in the U.S. came from just 20% of institutions [@wapmanQuantifyingHierarchyDynamics2022a]. Within anthropology, such studies draw attention to a central paradox—a discipline purportedly committed to fighting social inequalities continues to reproduce its own systemic inequalities through hiring practices that favor an elite minority of candidates with prestigious affiliations.

@@

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. Although some time has passed since the UC experiment, institutional attitudes to diversity are likely to remain in flux for the foreseeable future. The 2023 Supreme Court decision in Students for Fair Admissions v. Harvard, which effectively ended affirmative action at American colleges and universities, in tandem with the Trump administration’s 2025 targeting of DEI initiatives [@knox2025] foreshadow a shift in judicial and political currents that is reshaping institutional approaches to—and even definitions of—diversity. The 180° pivot over the past five years—from requiring carefully crafted diversity statements as a matter of course to abrupt and sweeping policy shifts rendering such documents obsolete—indicates that national and institutional commitments to diversity may be only as deep as the latest strategic plan.

@@

Collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of editorial activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. The MediaWiki technology used by Academic Jobs Wiki creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics, which allows us to measure the dynamics of editing activity over the ten years of our study period. Every edit includes the author's identity, recorded as either an IP address (a numerical address that includes information about the location of the user's computer) or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the U.S., indicating the Archaeology Jobs Wiki is primarily edited and visited by users in the U.S. The number of edits and active editors roughly halved from 2013 to 2017, and then continued to gradually decline. Despite this downward trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant across our study period. The majority of editors are active on the site for a single day (n = `r edits_for_each_year_tbl_span_tbl_less_than_one_day`, `r edits_for_each_year_tbl_span_tbl_less_than_one_day_perc`%). There are `r edits_for_each_year_tbl_span_tbl_less_than_one_year` (`r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%) editors who were active for between one day and one year. Only `r edits_for_each_year_tbl_span_tbl_more_than_one_year` (n = `r edits_for_each_year_tbl_span_tbl_more_than_one_year_perc`%) editors were active for more than one year, with `r max(round(edits_for_each_year_tbl_span_tbl$edit_span_years,0))` years as the longest period any single editor was active on the Wiki. For those active for more than one day, the average duration of editorial activity is `r edits_for_each_year_tbl_span_tbl_mean` years. This suggests that most of the editing activity on the Wiki reflects a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads. That said, inferring the number of editors from the number of IP addresses is complicated by the high mobility of job seekers. For example one person could be represented by several IP addresses as they relocate from one city to another while working in a series of short-term appointments such as post-doctoral fellowships and adjunct teaching positions. This might result in an overestimation of the number of editors on the Wiki and underestimation of the duration of their editorial activity.

@@

This trajectory in archaeology—where there appears to be a temporal "sweet spot" post-PhD for obtaining permanent academic positions—dovetails with a larger meta-analysis of the career trajectories of PhDs from the humanities and humanistic social sciences. @main2019 used longitudinal Graduate Education Survey (GES) data from the Andrew W. Mellon Foundation to analyse the career trajectories of over 5,000 PhDs from disciplines in the humanities and humanistic social sciences at three time points: six months post PhD, three years post PhD, and eight years post-PhD. The proportion of surveyed doctorates in tenure-track or tenured faculty positions increased with time since PhD—41% of the sample held a tenure-track position six months after defending, 64% held a tenure-track position three years after defending, and 68% held a tenure-track or tenured position eight years after defending. In keeping with broader patterns observed across the humanities and social sciences, the variable tenures of editors on the Archaeology Jobs Wiki may partially reflect the temporal patterns in hiring relative to job seekers' time since PhD.

There seems to be a shift over time in the balance of editors who identify as job-seekers versus search committee members, as indicated by the number of self-reporting users over time. Panel G of @fig-panel-wiki-characteristics shows the data found on each page in a section labeled 'current users', where users can volunteer to update a table tracking the status of users. These data indicate that the total number of self-reporting users, from the values recording at 'How many people use the wiki?', declined by `r how_many_people_use_the_wiki_perc_start_end`% during the study period. At the same time, the proportion of users self-identifying as "search committee members" increased, from under `r search_committee_members_perc_start`% in 2014–2018 to over `r search_committee_members_perc_end`% from 2020 onward. This suggests that, over time, fewer job applicants have been contributing job postings and sharing information about their experiences, and search committee members have increasingly been posting job advertisements themselves. Overall, the Archaeology Academic Jobs Wiki is best understood not as a comprehensive and neutral archive of all jobs posted during the study period, but a biased sample of the the jobs that early career applicants were most focused on applying for, and sharing information about.

@@

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013–2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methodological foci of the position into a Google form. The topical focus is what we understood as the intellectual core of the position—examples of topical foci included environmental archaeology, public archaeology, and North American archaeology. The geographic focus is the region of the world about which the ideal candidate has scholarly expertise, for example, Southwest US, Mediterranean, or Asia and India. The methods focus is the data-generating sub-field of archaeology mentioned in the ad. Examples of methods used in this study include archaeobotany, lithic analysis, and zooarchaeology. We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad.

After completing primary data collection, we studied the topical, geographic, and methods text of each ad. Following the approach of @ryan2003techniques, we collaboratively and manually reduced the variation in the raw data into 10–15 categories appearing in at least 20 (for topics and geography) or 10 (for methods) job ads to simplify analysis and visualization. This means that some topics, such as gender (mentioned in `r n_jobs_mentioning_gender` ads) do not appear in our results because of their rarity in the job ads. Full details of the category reduction, showing the mapping between exact phrases found in the job ads and the categories we used for our analysis, can be found in our Supplementary Materials. Our final topic categories were: American archaeology, Ancient Europe and Mediterranean, Archaeological science, Archaeological theory, Bioarchaeology, Complex societies, Digital archaeology, Environmental archaeology, Evolutionary anthropology, Indigenous and historical archaeology, North Mesoamerican archaeology, Pleistocene archaeology, and Public archaeology. Our geographic categories were: Africa, Americas, Asia & India, Canada & Arctic, Europe, Mediterranean, Meso- & South America, Near East, Oceania, Midwest US, Northeastern US, Southeast US, Southwest US, and Western US. Our methods categories were: Archaeobotany, Archaeometry, Bioarchaeology, Ceramic analysis, Computational and Digital archaeology, Geoarchaeology, Landscape analysis, Lithic analysis, Material culture analysis, and Zooarchaeology.

@@

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/10.5281/zenodo.14798941 to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013–2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at U.S. universities. @fig-show-basic-plots shows the count of ads for each year, where year refers to the year the job ad was posted. @tbl-show-basic-counts shows the breakdown by different job types, ranks, and tenure status. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions. Only academic year 2013–2014 had more non-tenure-track positions than tenure-track positions, which was followed by an upward trend peaking at 2018–2019 and then declining again into the present.

@@

Panel A of @fig-show-map-of-hiring-institution-and-cc shows the frequencies of hiring institutions according to their Carnegie Classification, a framework for classifying U.S. colleges and universities according to the types of degrees awarded, levels of research activity, and topical foci [@shulman2001carnegie]. Doctoral universities with high and very high research activity are by far the most active institutions hiring archaeology faculty. Associate's colleges, also known as community colleges, rarely post job ads for archaeology faculty.

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013–2023 include New York, Texas, Pennsylvania, and Florida. These top five states correspond to the top five most populous U.S. states, suggesting that rates of hiring are approximately proportional to population density. The top five states for job ads are also the five states with highest number of degrees awarded in Anthropology [@IPEDSNCES]. Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period. The implication here is that job-seekers who are able to relocate to populous areas will have more employment options.

@@

We recorded all geographic regions mentioned in the text where the successful applicant should have expertise and be research active. Our analysis focuses on those locations mentioned in 20 or more ads. Overall, American locations dominate. Panel A of @fig-show-geo-trends shows that a single region of the US, the Southwest, occurs in more job ads than every other part of the world except for the Mediterranean. The Southwest includes Arizona and New Mexico, with portions of California, Colorado, Nevada, Oklahoma, Texas, and Utah. It is archaeologically significant as the home of the Ancestral Pueblo, Hohokam, and Mogollon peoples, who practiced irrigation agriculture and lived in relatively large settlements compared to other regions of the U.S. The area was later occupied by the Navajo, Ute, Southern Paiute, Hopi and Zuni, groups who had similarly high population densities [@griffin2000native]. The Mediterranean is prominent because it is the region that is often mentioned in job ads focused on classical archaeology (i.e. archaeology of Bronze Age and Iron Age Italy and Greece).
